Rhizaria in the oligotrophic ocean exhibit clear temporal and vertical variability.

### 2024-01-29 manuscript/03_results.qmd
render with line numbers

![Environmental profiles across time-series of study period. Y axis shows depth in meters. (A) Temperature. (B) Salinity. (C) Dissolved Oxygen. (D) In-situ chlorophyll fluorescence. (E) Particle concenctrion (184 - 450$\mu m$). (F) Bacteria Abundance. (G) Silica. (H) Nitrate](images/02-environmental.pdf)

Overall, particle concentration was high near the surface during the 2021 spring bloom, then sank throughout the water column attenuating throughout the lower epipelagic (Figure 2E). Similarly, bacteria abundance was closely linked to overall productivity, althrough there was a more consistent layer near the top of the mesopelagic (\~250m) (Figure 2F). Concurrent with the secondary fall production peak, there was also higher particle concentration and bacterial abundance in the later summer and early fall. Interestingly, while primary productivity estimates from July-August were not that different between 2019 and 2021 (Figure 4A), chlorophyll-a florescence, particle concentration, and bacterial abundance were much higher in 2019's summer/fall (Figure 2D-F). Inorganic nutrients ($Si$ and $NO_3$) were generally well stratified, with low concentrations in the epipelagic and increasing throughout the mesopelagic. However, both nutrients did vary vertically in accordance with the 2021 February downwelling and spring mixing period (Figure 2G-H). Additionally, in the late fall of 2021, $Si$ concentrations were slightly elevated in the mid-mesopelagic (Figure 2G).

@@

![Primary productivity (A) and flux estimates from total mass (B), carbon (C), and nitrogen (D). Values are from monthly cruises with month displayed in corresponding colors. Absent data are shown by grey bar.](images/03_prod-flux.pdf)

@@

Across all imaged mesozooplankton (\>900$\mu m$), Rhizaria comprised a considerable fraction of the total community. Considering the total abundances of the observational period, Rhizaria comprised on average, 42.6% of all mesozooplankton abundance (Supplemental Figure 3). Copepods were the second most abundant, comprising 35.5% and all living mesozooplankton were 22%. The large contribution of Rhizaria to the mesozooplankton community is most prominent in the epipelagic, where they accounted for 47% of all mesozooplankton. In the mesopelagic rhizaria were a smaller fraction, at 38% in the upper layers (200-500m) and 37% in the deeper mesopelagic (500-1000m).

![Average abundance of Rhizaria in 25m bins, across entire study period. Shown are total Rhizaria (A), Acantharea (B), Collodaria (C), Aulacanthidae (D), Foraminifera (E), Aulosphaeridae (F), Castanellidae (G), Coelodendridae (H).](images/04_vertical.pdf)

Total average Rhizaria abundance had a bimodal distribution with respect to depth. Total abundance was highest just below the surface (0-100m), with secondary, wider peak occurring in the mid mesopelagic (Figure 4A). Although, variation in depth binned abundance was large, likely due to seasonal variability but also increased from the detection-risk described in the methods. The vertical distribution pattern and abundance varied considerably across taxonomic groups. Radiolarians were some fo the most abundant taxa observed, particularly in the epipelagic (Figure 4, Figure 5B). This pattern was led by Collodaria, whose colonies were abundant in the upper epipelagic and declined into the top of the mesopelagic (Figure 4C). Acantharea displayed a bimodal distribution accounting for a large portion of the total Rhizaria pattern (Figure 4B, Figure 5). Foraminifera had a similar bimodal distribution, yet their overall average densities were much lower and spread wider throughout the mesopelagic (FIgure 4E). Phaeodarian families exhibited a wide range of vertical distribution patterns. The most abundant, Aulacanthidae, also had a bimodal pattern but the density was highest in the lower mesopelagic (Figure 4D). Aulosphaeridae had low average density and was nearly homogenously distributed throughout the water column, although slightly lower int eh epipelagic (Figure 4F). Castanellidae were the only Phaeodarian who appeared to be effectively restricted to the photic zone (Figure 4G). Alternatively, Coelodendridae primarily occurred in the lower mesopelagic (Figure 4H). A few individuals from the families Tuscaroridae and Medusettidae were also observed in the mesopelagic, yet they were much rarer (data not shown).

@@

![Seasonality of Rhizarian integrated abundance for the epipelagic (0-200m) (A-B), upper mesopelagic (200-500m) (C-D), lower mesopelagic (500-1000m) (E-F). Left panels (A,C,E) display total integrated abundance per monthly cruise colored by month. Right panels (B, D, F) display community composition of each total abundance.](images/05_seasonality.pdf)

@@

| **Model**                          | Term                   | edf    | F     | p-value |
|------------------------------------|------------------------|--------|-------|---------|
| **All Rhizaria Epipelagic**        | Salinity               | 2.0818 | 4.654 | \<0.001 |
| $R^2_{adj}$= 0.42                  | DO                     | 0.8272 | 0.945 | 0.0136  |
|                                    | Silica                 | 3.1206 | 15.75 | \<0.001 |
|                                    | NO3                    | 2.9970 | 6.268 | \<0.001 |
|                                    | Primary Productivity   | 1.8367 | 2.099 | \<0.001 |
|                                    | Particle Concentration | 0.9282 | 2.566 | \<0.001 |
|                                    |                        |        |       |         |
| **All Rhizaria Upper Mesopelagic** | Silica                 | 0.6847 | 0.431 | 0.073   |
| $R^2_{adj}$= 0.406                 | Avg Mass Flux          | 1.5426 | 0.923 | 0.045   |
|                                    | Particle Concentration | 3.2957 | 20.89 | \<0.001 |
|                                    |                        |        |       |         |
| **All Rhizaria Lower Mesopelagic** | Avg Mass Flux          | 1.6632 | 1.824 | 0.002   |
| $R^2_{adj}$= 0.603                 | Particle Concentration | 0.7694 | 0.662 | 0.027   |

: Generalized Additive Model results for integrated total Rhizarian abundance in different regions of the water column.

GAMs for individual taxa were much less consistent in their fits (Table 2). This is likely in part due to the high number of non-observations for certain taxa. Note that due to low abundances, GAMS were not constructed for Tuscaroridae or Medusettidae. Furthermore no significant terms were found for a model with Aulosphaeridae in the epipelagic nor Foraminifera in the mesopelagic.

![Partial effects of smooth terms in taxa-specific GAM models from the epipelagic (0-200m). Effects are grouped by taxa; Acantharea (A), Foraminifera (B), Aulacanthidae (C), Collodaria (D), Castanellidae (E).](images/06_epi-partials.pdf)

Epipelagic Acantharea were explained by several predictor vairables and had a good fit ($R^2_{adj}$ = 0.53, Table 2). Most notable smooths however, were mass flux and particle concentration, which had a weak positive association (Figure 6A), with July 2021 as a clear outlier where Acantharean abundances were high in the epipelagic despite lower fluxes and particle concentrations (Figure 5). Foraminifera had a good fitting GAM in the epipelagic ($R^2_{adj}$ = 0.445). There were several significant explanatory variables, althoguht the clearest pattern was observed of high temperatures associated with more Foraminifera abundance (Table 2, Figure 6B). Epipelagic Aulacanthidae similarly had several predictor variables which were significant, including both water quality parameters and particle/flux predictors (Table 2). Interestingly, Aulacanthidae had primary production as a significant predictor, yet there was not a clear association (Figure 6C). There was a fit for Collodaria in the epipelagic ($R^2_{adj}$ = 0.16), although there was a logit-like relationship where higher abundances tended to occur during higher DO conditions in the surface waters (Figure 6D). Castanellidae also had similarly poor fits in the epipelagic ($R^2_{adj}$ = 0.124) (Table 2, Figure 6E).

| Model                                | Term                   | edf   | F     | p-value |
|--------------------------------------|------------------------|-------|-------|---------|
| **Acantharea Epipelagic**            | Salinity               | 2.264 | 4.113 | \<0.001 |
| R2adj=0.53                           | O2                     | 2.579 | 6.712 | \<0.001 |
|                                      | Avg Mass Flux          | 4.810 | 22.81 | \<0.001 |
|                                      | Bacteria #/L           | 1.599 | 1.317 | 0.0117  |
|                                      | Particle Concentration | 0.853 | 1.158 | 0.0087  |
|                                      |                        |       |       |         |
| **Acantharea Upper Mesopelagic**     | Avg C Flux             | 1.296 | 0.621 | 0.0439  |
| R2adj=0.231                          | Avg N Flux             | 1.619 | 1.283 | 0.0026  |
|                                      | Particle Concentration | 0.952 | 3.886 | \<0.001 |
|                                      |                        |       |       |         |
| **Acantharea Lower Mesopelagic**     | Temperature            | 2.076 | 4.155 | \<0.001 |
| R2adj=0.509                          | Avg N Flux             | 1.494 | 1.216 | 0.0113  |
|                                      | Primary Productivity   | 0.766 | 0.648 | 0.0253  |
|                                      | Particle Concentration | 2.037 | 10.03 | \<0.001 |
| **Aulacanthidae Epipelagic**         | Salinity               | 0.792 | 0.757 | 0.0241  |
| R2adj=0.251                          | Avg N Flux             | 0.869 | 1.324 | 0.0018  |
|                                      | Primary Productivity   | 2.312 | 7.704 | \<0.001 |
|                                      | Particle Concentration | 2.008 | 2.200 | 0.0017  |
|                                      |                        |       |       |         |
| **Aulacanthidae Upper Mesopelagic**  | Particle Concentration | 2.832 | 9.802 | \<0.001 |
| R2adj = 0.158                        |                        |       |       |         |
|                                      |                        |       |       |         |
| **Aulacanthidae Lower Mesopelagic**  | Avg Mass Flux          | 1.622 | 1.991 | 0.002   |
| R2adj=0.298                          | Particle Concentration | 2.123 | 6.706 | \<0.001 |
|                                      |                        |       |       |         |
| **Aulosphaeridae Upper Mesopelagic** | Particle Concentration | 2.653 | 13.06 | \<0.001 |
| R2adj=0.2                            |                        |       |       |         |
|                                      |                        |       |       |         |
| **Aulosphaeridae Lower Mesopelagic** | Particle Concentration | 0.972 | 6.248 | \<0.001 |
| R2adj=.147                           |                        |       |       |         |
|                                      |                        |       |       |         |
| **Castanellidae Epipelagic**         | Avg C Flux             | 1.462 | 0.816 | 0.0421  |
| R2adj = 0.124                        | Primary Productivity   | 0.771 | 0.665 | 0.0247  |
|                                      | Particle Concentration | 3.956 | 4.623 | \<0.001 |
|                                      |                        |       |       |         |
| **Coelodendridae Upper Mesopelagic** | Avg N Flux             | 0.822 | 0.919 | 0.0183  |
| R2adj=.113                           | Particle Concentration | 0.970 | 6.208 | \<0.001 |
|                                      |                        |       |       |         |
| **Coelodendridae Lower Mesopelagic** | Particle Concentration | 1.773 | 4.873 | \<0.001 |
| R2adj=.133                           |                        |       |       |         |
|                                      |                        |       |       |         |
| **Collodaria Epipelagic**            | Salinity               | 0.925 | 2.267 | \<0.001 |
| R2adj=0.16                           | O2                     | 2.015 | 2.217 | 0.002   |
|                                      | Bacteria #/L           | 1.843 | 2.100 | 0.002   |
|                                      |                        |       |       |         |
| **Foraminifera Epipelagic**          | Temperature            | 4.414 | 8.789 | \<0.001 |
| R2adj=0.445                          | O2                     | 1.603 | 1.287 | 0.0111  |
|                                      | Avg N Flux             | 2.512 | 3.396 | \<0.001 |
|                                      | Primary Productivity   | 0.824 | 0.920 | 0.0153  |
|                                      | Particle Concentration | 0.919 | 2.257 | \<0.001 |

: Taxa-specific generalized additive models for different regions of the water column.

![Partial effects of smooth terms in taxa-specific GAM models from the upper mesopelagic (200-500m). Effects are grouped by taxa; Acantharea (A), Aulacanthidae (B), Aulosphaeridae (C), Coelodendridae (D).](images/07_upmeso-partial.pdf)

In the upper mesopelagic (200-500m), abundances were generally low (Figure 4) so GAMs were only constructed for Acantharea, Aulacanthidae, Aulosphaeridae, and Coelodendridae (Table 2). All these models had generally poor fits ($R^2_{adj}$ \< 0.25). Yet, for all upper mesopelagic models, particle concentration was a significant explanatory variable (Table 2, Figure 7). Carbon flux was significant for Acantharea and nitrogen flux was significant for both Acantharea and Coelodendridae (Table 2, Figure 7A,D). The lower mesopelagic also had generally poor GAM fits for taxa specific models ($R^2_{adj}$ \<0.3), with the exception of Acantharea ($R^2_{adj}$ = 0.509). Acantharea in the lower mesopelagic was most clearly positively associated with particle concentration and nitrogen flux, as well as temperature to a slight degree (Figure 8A). For all Phaeodarians with a significant model, particle concentration was a main predictor variable (Table 2, Figure 8B-D). Aulacanthidae had the best fitting model of the Phaeodarians ($R^2_{adj}$ = 0.298), which also included mass flux as a statistically significant smooth (Figure 8B).

![Partial effects of smooth terms in taxa-specific GAM models from the lower mesopelagic (500-1000m). Effects are grouped by taxa; Acantharea (A), Aulacanthidae (B), Aulosphaeridae (C), Coelodendridae (D).](images/08_lomeso-partials.pdf)

>>>>

Overall, particle concentration was high near the surface during the 2021 spring bloom, then moved deeper throughout the water column attenuating throughout the lower epipelagic (Figure 2E). Similarly, bacteria abundance was closely linked to overall productivity, althrough there was a more consistent moderate-abundance layer near the top of the mesopelagic (\~250m) (Figure 2F). Concurrent with the secondary fall production peak, there was also higher particle concentration and bacterial abundance in the later summer and early fall. Interestingly, while primary productivity estimates from July-August were not that different between 2019 and 2021 (Figure 4A), chlorophyll-a florescence, particle concentration, and bacterial abundance were much higher in 2019's summer/fall (Figure 2D-F). Inorganic nutrients ($Si$ and $NO_3$) were generally well stratified, with low concentrations in the epipelagic and increasing throughout the mesopelagic. However, both nutrients did vary vertically in accordance with the 2021 February downwelling and spring mixing period (Figure 2G-H). Additionally, in the late fall of 2021, $Si$ concentrations were slightly elevated in the mid-mesopelagic (Figure 2G).

@@

@@

Across all imaged mesozooplankton (\>900$\mu m$), Rhizaria comprised a considerable fraction of the total community. Considering the total abundances of the observational period, Rhizaria comprised on average, 42.6% of all mesozooplankton abundance (Supplemental Figure 3). Copepods were the second most abundant, comprising 35.5% and all other living mesozooplankton were 22%. The large contribution of Rhizaria to the mesozooplankton community is most prominent in the epipelagic, where they accounted for 47% of all mesozooplankton. In the mesopelagic rhizaria were a smaller fraction, at 38% in the upper layers (200-500m) and 37% in the deeper mesopelagic (500-1000m).

Total average Rhizaria abundance had a bimodal distribution with respect to depth. Total abundance was highest just below the surface (0-100m), with secondary, wider peak occurring in the mid mesopelagic (Figure 4A). Variation in depth binned abundance was large, likely due to seasonal variability but also increased from the detection-risk described in the methods. The vertical distribution pattern and abundance varied considerably across taxonomic groups. Radiolarians were some of the most abundant taxa observed, particularly in the epipelagic (Figure 4, Figure 5B). This pattern was led by Collodaria, whose colonies were abundant in the upper epipelagic and declined into the top of the mesopelagic (Figure 4C). Acantharea displayed a bimodal distribution accounting for a large portion of the total Rhizaria pattern (Figure 4B, Figure 5). Foraminifera had a similar bimodal distribution, yet their overall average densities were much lower and spread wider throughout the mesopelagic (Figure 4E). Phaeodarian families exhibited a wide range of vertical distribution patterns. The most abundant, Aulacanthidae, also had a bimodal pattern but the density was highest in the lower mesopelagic (Figure 4D). Aulosphaeridae had low average density and was nearly homogeneously distributed throughout the water column, although slightly lower in the epipelagic (Figure 4F). Castanellidae were the only Phaeodarian who appeared to be effectively restricted to the photic zone (Figure 4G). Alternatively, Coelodendridae primarily occurred in the lower mesopelagic (Figure 4H). A few individuals from the families Tuscaroridae and Medusettidae were also observed in the mesopelagic, yet they were much rarer (data not shown).

@@

### Body size throughout the water column.

Very few taxa had consistent distributions throughout the water column. Only Acanthrea, Foraminifera, Aulacanthidae, and Aulosphaeridae were consistently abundant in the epipelagic and mesopelagic. To investigate if the populations or morphologies shifted throughout the water column, we compared the sizes (Equivalent Spherical Diameters, ESD) between mesopelagic and epipelagic groups for each taxa. All groups were significantly different on average (Wilcox Rank Sum p-value <0.001). Acantharea were smaller, on average in the mesopelagic while all other taxa tended to be larger (Figure 6).

@@

GAMs for individual taxa were much less consistent in their fits (Table 2). This is likely in part due to the high number of non-observations for certain taxa. Note that due to low abundances, GAMs were not constructed for Tuscaroridae or Medusettidae. Furthermore no significant terms were found for a model with Aulosphaeridae in the epipelagic nor Foraminifera in the mesopelagic.

Epipelagic Acantharea were explained by several predictor variables and had a good fit ($R^2_{adj}$ = 0.53, Table 2). Most notable smooths were mass flux and particle concentration, which had a weak positive association (Figure 7A), with July 2021 as a clear outlier where Acantharean abundances were high in the epipelagic despite lower fluxes and particle concentrations (Figure 5). Foraminifera had a good fitting GAM in the epipelagic ($R^2_{adj}$ = 0.445). There were several significant explanatory variables, although the clearest pattern was observed of high temperatures associated with more Foraminifera abundance (Table 2, Figure 7B). Epipelagic Aulacanthidae similarly had several predictor variables which were significant, including both water quality parameters and particle/flux predictors (Table 2). Interestingly, Aulacanthidae had primary production as a significant predictor, yet there was not a clear association (Figure 7C). There was a fit for Collodaria in the epipelagic ($R^2_{adj}$ = 0.16), although there was a logit-like relationship where higher abundances tended to occur during higher DO conditions in the surface waters (Figure 7D). Castanellidae also had similarly poor fits in the epipelagic ($R^2_{adj}$ = 0.124) (Table 2, Figure 7E).

In the upper mesopelagic (200-500m), abundances were generally low (Figure 4) so GAMs were only constructed for Acantharea, Aulacanthidae, Aulosphaeridae, and Coelodendridae (Table 2). All these models had generally poor fits ($R^2_{adj}$ \< 0.25). Yet, for all upper mesopelagic models, particle concentration was a significant explanatory variable (Table 2, Figure 8). Carbon flux was significant for Acantharea and nitrogen flux was significant for both Acantharea and Coelodendridae (Table 2, Figure 8A,D). The lower mesopelagic also had generally poor GAM fits for taxa specific models ($R^2_{adj}$ \<0.3), with the exception of Acantharea ($R^2_{adj}$ = 0.509). Acantharea in the lower mesopelagic was most clearly positively associated with particle concentration and nitrogen flux, as well as temperature to a slight degree (Figure 9A). For all Phaeodarians with a significant model, particle concentration was a main predictor variable (Table 2, Figure 9B-D). Aulacanthidae had the best fitting model of the Phaeodarians ($R^2_{adj}$ = 0.298), which also included mass flux as a statistically significant smooth (Figure 9B).

### 2024-01-29 manuscript/04_discussion.qmd
render with line numbers

Overall Rhizaria were the most commonly identified group of mesozooplankton throughout the study period. We do note that the UVP5 commonly captures Trichodesmium colonies, yet these were excluded in this comparison as they are strictly autotrophs. It should be noted that previous work has suggested that avoidance behavior with the UVP is possible, at times likely, for visual and highly mobile zooplankton [@barth2022]. Thus, the percent contribution reported here (42.7%) of Rhizaria to the total mesozooplankton community may be inflated due to under sampling of organisms such as Euphausiids and Chaetognaths which has quick escape responses. Regardless, it is worth noting that in the same region, with data collected in 2012 and 2013 using similar calculation methods, @biard2016 estimated Rhizaria only contribute 15% of the total mesozooplankton community in the upper 500m. Likely, Rhizaria display considerable interannual variability. In the present study, we noticed considerably higher Rhizarian abundance throughout the water column in 2019 compared to 2021. While this may have been driven by increased mass flux (Figure 3), more information is needed to truly understand the magnitude by which Rhizaria can vary interannually.

@@

In this study we present a clear pattern of vertical zonation between different Rhizaria groups. Largely, the taxonomic composition and vertical positioning were similar to Rhizaria zonation in the California Current Ecosystem [@biard2020]. It should be noted however, that the secondary abundance peak reported in the present study is lower. This is likely due to the more oligotrophic nature of the study region, were the euphotic zone penetrates deeper into the water column. Most prevalent in the epipelagic were Collodaria. These mixotrophic Radiolarians have long been reported to contribute to primary productivity in the euphotic zone [@michaels1995; @dennett2002]. Collodaria are thought to be particularly successful globally in oligotrophic regions due to their photosymbiotic relationships [@biard2016; @biard2017]. We observed the highest abundance of Collodaria during June 2021, supporting the notion they can thrive during the typically low-nutrient conditions of summer stratification. However, Collodaria also increased during the spring mixing period, suggesting that they can thrive during conditions which may typically be thought to favor autotrophs. Furthermore, while Collodaria were primarily absent from below 250m, there were a few instances of deeper colonies being observed. Global investigations of polycistine flux, suggest that deep-Collodaria in Oligotrophic regions may be a consequence of isothermal submersion [@boltovskoy2017]. Another effectively exclusively epipelagic Rhizaria was the Phaeodarian family of Castanellidae. All Phaeodarians are thought to be fully heterotrophic [@nakamura2015], nonetheless a number of studies, including this one, report Castanellidae to be typically found in the lower epipelagic [@zasko2005; @biard2018; @biard2020]. It should be considered that perhaps Castanellidae specializes in feeding sinking particles directly at the base of the epipelagic. Given it's smaller size [@nakamura2015], Castanellidae does not need a large diameter to efficiently flux feed at the typically particle rich region of the lower epipelagic. Both Castanelldiae and Collodaria had poor fitting GAMs. This is somewhat of a surprise for Collodaria who had large abundance. However, given the consistency of their abundance, it may be that this study did not capture a wide enough range of conditions for describing Collodaria's preferred niche.

@@

Acantharea also had a bimodal distribution, with much larger abundances than Aulacanthidae or Foraminifera. Most prior studies of Acantharea vertical distribtion found them concentrated in near surface layers of the water column [@michaels1988 @zasko2005; @biard2020]. This would support the paradigm that large Acantharean abundances may be supported by their mixotrophic abilities [@michaels1995; @suzuki2015]. While the UVP5 images cannot distinguish between mixotrophic and heterotrophic Acantharea, the GAMs constructed for Acantharean abundance found positive associations with particle concentration and mass flux, suggesting a higher reliance on heterotrophy. Recently @marsbrisbin2020 described apparent predator behavior amongst near-surface Acantharea. Thus it is likely that epipelagic Acantharea may commonly be heterotrophic. Yet, it should be noted in the Sargasso Sea, both heterotrophic and symbiotic lineages of Acantharea have been reported [@blanco-bercial2022]. Additionally, @michaels1988 noted that the majority of Acantharea (by abundance) were smaller than 160$\mu m$. While that estimate may be inflated due to inability to capture larger cells, small Acantharea were not captured in the present study. Thus, trophic strategy may shift based on sizes of Acantharea.

@decelle2013 proposed a hypothetical life cycle for cyst-bearing (strictly heterotrophic) Acantharea. This hypothesized life cycle suggests that epipelagic Acantharea are adult populations, which form cysts that sink into the mesopelagic, then reproduce and raise. Furthermore, given that horizontal transfer of symbionts between generations of Acantharea is unlikely, the newly spawned mesopelagic Acantharea are not necessarily required to rapidly return to the photic zone. This hypothesis predicts that Acanthareans in the mesopelagic would be smaller [@decelle2013]. @marsbrisbin2020 provided some support for this hypothesis, with a significant decrease in Acantharea sizes with depth. Although, the authors also observed low abundances in the mesopelagic and noted that the smaller sizes may be due to less food availability [@marsbrisbin2020]. Since food is more scarce in the mesopelagic there is is less nutritional quality, yet flux feeders would likely grow larger to increase their feeding range [@biard2020]. In the data collecting in this study, Acanthareans in the mesopelagic were significantly smaller than the epipelagic, despite the other bimodal taxa (Foraminifera and Aulacanthidae) being significantly larger with depth (Figure 9, Wilcoxon rank sum text p-value \< 0.001). This provides added support for the hypothesis that cyst-forming Acantharea may utilize different sections of the water column throughout their life cycle. However to further investigate this, more work is needed with higher temporal and taxonomic resolution.

![Comparison of average sizes (ESD) amongst Rhizarian taxa which occurred throughout the water column.](images/09_size-comp.pdf)

@@

This study provides a detailed look at Rhizarian abundance over time throughout the water column in a major oligotrophic gyre. We show that their abundances are generally related to particle concentration and flux, although lack of environmental variability may have reduced the fit of our GAMs. Considering the potential role of Rhizaria in the biological carbon pump, they may have a somewhat mixed role. In the shallower regions, where smaller Rhizaria are abundant, they may be an attenuating force on sinking particles [@stukel2019]. It should be noted that in our study, we focused on the "small" particle concentration field (\<450$\mu m$), and these particles are generally slower sinking than large particles. However, once consumed and repackaged by larger Rhizaria, they can sink quicker and contribute more to overall flux [@michaels1988]. Thus, Rhizaria may act as a ballasting mechanism. However, this is largely speculation, to truly test this, more work is needed measuring Rhizarian flux.

The vertical partitioning documented in this study do support the hypothesis that mixotrophic rhizaria will occupy shallower waters while deeper waters are dominated by heterotrophy. However the degree to which mixotrophic Rhizaria in the euphotic zone rely on heterotrophy versus symbiosis is uncertain. Collodaria were recorded as consistent and dominant members of the near surface region. These organisms have the potential to contribute considerably to the otherwise low productivity of oligotrophic regions. However, their role in food webs is not well understood. While this study represents a step-forward in our understanding of Rhizarian ecology, there are several known aspects of Rhizarian and protist ecology which are not captured in this study. Continued research on Rhizaria is much needed to better understand their ecology. Particularly extended descriptive work to capture interannual patterns. Also work defining biotic interactions, feeding rates, productivity, and life history are all rich fields of interest in Rhizaria.

>>>>

Overall Rhizaria were the most commonly identified group of mesozooplankton throughout the study period. We do note that the UVP5 commonly captures Trichodesmium colonies, yet these were excluded in this comparison as they are strictly autotrophs. It should be noted that previous work has suggested that avoidance behavior with the UVP is possible, at times likely, for visual and highly mobile zooplankton [@barth2022]. Thus, the percent contribution reported here (42.7%) of Rhizaria to the total mesozooplankton community may be inflated due to under sampling of organisms such as Euphausiids and Chaetognaths which have quick escape responses. Regardless, it is worth noting that in the same region, with data collected in 2012 and 2013 using similar calculation methods, @biard2016 estimated Rhizaria only contribute 15% of the total mesozooplankton community in the upper 500m. Likely, Rhizaria display considerable interannual variability. In the present study, we noticed considerably higher Rhizarian abundance throughout the water column in 2019 compared to 2021. While this may have been driven by increased mass flux, more information is needed to truly understand the magnitude by which Rhizaria can vary interannually.

### Relationship to environmental parameters

In general, the fit of most GAMs were moderate to poor. One possible reason for the poor fits may have been that for some taxa, conditions were not variable enough to capture a range of conditions at which they may exist. For instance, Collodaria were the most abundant taxa observed, yet the fit of their GAM was particularly poor. In studies which covered a wider range of parameters, Collodaria has been shown to strongly vary with changes in parameters such as temperature, chlorophyll-a, mixing, and water clarity [@biard2017; @biard2020]. Alternatively, Acantharea had relatively good fitting GAMs. These taxa also had some of the largest variation from month to month on cruises. Thus, it may be that in the oligotrophic, the relatively stable conditions can support certain taxa while others are more sporadic. It should also be noted that due to the challenge of adequately sampling enough volume to overcome low-detection issues, GAMs were run on integrated data. However, variation with environmental parameters throughout the water column are likely, just not captured in the modelling aspect of this study. One consistent parameter which had significant positive associations was particle concentration. This observation is not surprising as most rhizaria likely to some extent engage in flux feeding.

@@

In this study we present a clear pattern of vertical zonation between different Rhizaria groups. Largely, the taxonomic composition and vertical positioning were similar to Rhizaria zonation in the California Current Ecosystem [@biard2020]. It should be noted however, that the secondary abundance peak reported in the present study is lower. This is likely due to the more oligotrophic nature of the study region, were the euphotic zone penetrates deeper into the water column. Most prevalent in the epipelagic were Collodaria. These mixotrophic Radiolarians have long been reported to contribute to primary productivity in the euphotic zone [@michaels1995; @dennett2002]. Collodaria are thought to be particularly successful globally in oligotrophic regions due to their photosymbiotic relationships [@biard2016; @biard2017]. We observed the highest abundance of Collodaria during June 2021, supporting the notion they can thrive during the typically low-nutrient conditions of summer stratification. However, Collodaria also increased during the spring mixing period, suggesting that they can thrive during conditions which may typically be thought to favor autotrophs. Furthermore, while Collodaria were primarily absent from below 250m, there were a few instances of deeper colonies being observed. Global investigations of polycystine flux, suggest that deep-Collodaria in Oligotrophic regions may be a consequence of isothermal submersion [@boltovskoy2017]. Another effectively exclusively epipelagic Rhizaria was the Phaeodarian family of Castanellidae. All Phaeodarians are thought to be fully heterotrophic [@nakamura2015], nonetheless a number of studies, including this one, report Castanellidae to be typically found in the lower epipelagic [@zasko2005; @biard2018; @biard2020]. It should be considered that perhaps Castanellidae specializes in feeding on sinking particles directly at the base of the epipelagic. Given it's smaller size [@nakamura2015], Castanellidae does not need a large diameter to efficiently flux feed at the typically particle rich region of the lower epipelagic. Both Castanelldiae and Collodaria had poor fitting GAMs. This is somewhat of a surprise for Collodaria who had large abundance. However, given the consistency of their abundance, it may be that this study did not capture a wide enough range of conditions for describing Collodaria's preferred niche.

@@

Acantharea also had a bimodal distribution, with much larger abundances than Aulacanthidae or Foraminifera. Most prior studies of Acantharea vertical distribution found them concentrated in near surface layers of the water column [@michaels1988 @zasko2005; @biard2020]. This would support the paradigm that large Acantharean abundances may be supported by their mixotrophic abilities [@michaels1995; @suzuki2015]. While the UVP5 images cannot distinguish between mixotrophic and heterotrophic Acantharea, the GAMs constructed for Acantharean abundance found positive associations with particle concentration and mass flux, suggesting a higher reliance on heterotrophy. Recently @marsbrisbin2020 described apparent predator behavior amongst near-surface Acantharea. Thus it is likely that epipelagic Acantharea may commonly be heterotrophic. Yet, it should be noted in the Sargasso Sea, both heterotrophic and symbiotic lineages of Acantharea have been reported [@blanco-bercial2022]. Additionally, @michaels1988 noted that the majority of Acantharea (by abundance) were smaller than 160$\mu m$. While that estimate may be inflated due to inability to capture larger cells, small Acantharea were not captured in the present study. Thus, trophic strategy may shift based on sizes of Acantharea.

@decelle2013 proposed a hypothetical life cycle for cyst-bearing (strictly heterotrophic) Acantharea. This hypothesized life cycle suggests that epipelagic Acantharea are adult populations, which form cysts that sink into the mesopelagic, then reproduce and rise. Furthermore, given that horizontal transfer of symbionts between generations of Acantharea is unlikely due to their spawning behavior, the newly spawned mesopelagic Acantharea are not necessarily required to rapidly return to the photic zone [@decelle2012; @decelle2013]. This hypothesis predicts that Acanthareans in the mesopelagic would be smaller [@decelle2013]. @marsbrisbin2020 provided some support for this hypothesis, with a significant decrease in Acantharea sizes with depth. Although the authors also observed low abundances in the mesopelagic and noted that the smaller sizes may be due to lower food availability [@marsbrisbin2020]. Since food is more scarce in the mesopelagi, nutritional quality lower, yet flux feeders would likely grow larger to increase their feeding range [@biard2020]. In the data collecting in this study, Acanthareans in the mesopelagic were significantly smaller than the epipelagic, despite the other bimodal taxa (Foraminifera and Aulacanthidae) being significantly larger with depth. This provides added support for the hypothesis that cyst-forming Acantharea may utilize different sections of the water column throughout their life cycle. However to further investigate this, more work is needed with higher temporal and taxonomic resolution.

@@

This study provides a detailed look at Rhizarian abundance over time throughout the water column in a major oligotrophic gyre. We show that their abundances are generally related to particle concentration and flux, although lack of environmental variability may have reduced the fit of our GAMs. Considering the potential role of Rhizaria in the biological carbon pump, they may have a somewhat mixed role. In the shallower regions, where smaller Rhizaria are abundant, they may be an attenuating force on sinking particles [@stukel2019]. It should be noted that in our study, we focused on the "small" particle concentration field (\<450$\mu m$), and these particles are generally slower sinking than large particles. However, once consumed and repackaged by larger Rhizaria, they can sink quicker and contribute more to overall flux [@michaels1988]. Thus, Rhizaria may act as an aggregation mechanism. However, this is largely speculation, to truly test this, more work is needed measuring Rhizarian flux.

The vertical partitioning documented in this study do support the hypothesis that mixotrophic rhizaria will occupy shallower waters while deeper waters are dominated by heterotrophy. However the degree to which mixotrophic Rhizaria in the euphotic zone rely on heterotrophy versus symbiosis is uncertain. Collodaria were recorded as consistent and dominant members of the near surface region. These organisms have the potential to contribute considerably to the otherwise low productivity of oligotrophic regions. However, their role in food webs is not well understood. While this study represents a step-forward in our understanding of Rhizarian ecology, continued research on Rhizaria is much needed to better understand their ecology. Particularly extended descriptive work to capture interannual patterns. Also work defining biotic interactions, feeding rates, productivity, and life history are all rich fields of interest in Rhizaria.
